Explorative data structure comparison with application to \ldots

### 2017-02-28 article/article_v0.01.tex
Definition of hairplot has been changed. This has been described in the article, and implemented in the PCADSC package.
Beside this other editorial changes have been made.

\newcommand{\bfU}{\mbox{\boldmath$\mathsf{U}$}}
\newcommand{\bfu}{\mbox{\boldmath$\mathsf{u}$}}
\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
\newcommand{\RR}{\mathbb{R}}
\DeclareMathOperator*{\V}{V}
\newcommand{\argmax}{\text{argmax}}
\newcommand{\R}[1]{\texttt{#1}}

@@

%\title{Something that is not PCADSC}
\title{Structure comparison for multivariate data by\ldots / with application to \ldots}

@@

Though the idea sounds appealing, it is quite difficult to assess similarity of matrices, and moreover, this becomes increasinly difficult for large numbers of variables and thus high dimensional covariance matrices. There is simply too much information to consider at once.

@@

\emph{Principal component analysis} (PCA) may be interpreted in terms of the \emph{rank-q-reconstruction error}. To describe this consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, and let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ be their average. Suppose that we want to describe each of the observations by $q$ numbers instead of the original $d$ numbers. The rank-q-reconstruction error is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
\end{equation*} 
PCA ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error. Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of the empirical covariance $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^d$. Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ of $S$ in the columns. The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a sign whenever the eigenvalues are different. If some of the eigenvalue are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation. In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero. It is a result from linear algebra that the rank-q-reconstruction error for $q < d$ is achieved for
\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
\end{equation*}
and equals $\sum_{i=q+1}^d \lambda_i$. In the terminology of PCA the eigenvectors $\eta_j \in \RR^d$ are called \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}. The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called \emph{scores}. The $j$th loading can also found as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2
\end{align*}

@@

\subsection{PCA-based data structure comparison}
Above, we promised a method for intuitive, visual inspection of data structure similarities, but as of now, all intuition might have been lost in technicalities. The main point we want to emphasize from PCA is that we can interpret the eigenvector/eigenvalue-pairs of the covariance matrix as tools for obtaining a different representation of the dataset. Specifically, we can interpret the PCA loadings (eigenvectors) as weights that determine the relative influence of each variable in each component. If two different data sets with the same variables, but different samples of observations, then have similar loading patterns, the two data sets will agree on which variables explain the most variance and also \textit{how} this variance is explained.  Looking at the PCA loadings and the cumulative explained variance thus provides a straightforward non-parametric graphical approach for assessing similarity between two datasets. \\

@@

\begin{figure}
\center
\includegraphics[scale = 0.8]{essWallyPCADSC.pdf}
\caption{\hl{something}}
\label{plotESSPCADSCWally}
\end{figure}

@@

>>>>

\newcommand{\bfU}{\mbox{\boldmath$\mathsf{U}$}}
\newcommand{\bfu}{\mbox{\boldmath$\mathsf{u}$}}
\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
\newcommand{\RR}{\mathbb{R}}
\DeclareMathOperator*{\V}{V}
\newcommand{\argmax}{\text{argmax}}
\newcommand{\R}[1]{\texttt{#1}}
\newcommand{\acos}{\text{arccos}}

@@

%\title{Something that is not PCADSC}
%\title{Structure comparison for multivariate data by\ldots / with application to \ldots}
\title{Explorative data structure comparison with application to \ldots}

@@

Though the idea sounds appealing, it is quite difficult to assess similarity of matrices, and moreover, this becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices. There is simply too much information to consider at once.

@@

\emph{Principal component analysis} (PCA) may be interpreted in terms of the \emph{rank-q-reconstruction error}. To describe this consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, and let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ be their average. Suppose that we want to describe each of the observations by $q$ numbers instead of the original $d$ numbers. The rank-q-reconstruction error is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
\end{equation*} 
PCA ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error. Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of the empirical covariance $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^d$. Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ in the columns. The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different. If some of the eigenvalues are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation. In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero. It is a result from linear algebra that the rank-q-reconstruction error for $q < d$ is achieved for
\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
\end{equation*}
and equals $\sum_{j=q+1}^d \lambda_j$. In the terminology of PCA the eigenvectors $\eta_j \in \RR^d$ are called \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}. The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called \emph{scores}. The $j$th loading can also found as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2.
\end{align*}

@@

\subsection{PCA-based data structure comparison}
Above, we promised a method for intuitive, visual inspection of data structure similarities, but as of now, all intuition might have been lost in technicalities. The main point we want to emphasize from PCA is that we can interpret the eigenvector/eigenvalue-pairs of the covariance matrix as tools for obtaining a different representation of the dataset. Specifically, we can interpret the PCA loadings (eigenvectors) as weights that determine the relative influence of each variable in each component. If two different data sets with the same variables, but different samples of observations, then have similar loading patterns, the two data sets will agree on which variables explain the most variance and also \textit{how} this variance is explained.  Looking at the PCA loadings and the cumulative explained variance thus provides a straightforward non-parametric graphical approach for assessing similarity between two datasets.

{\color{red}
We propose the following three diagnostic plots:
\begin{description}
\item[CumVarAgreement:] Cumulated Bland-Altman plot for the variation decomposition.
\item[Hairplot:] Plot of the loadings for one of the datasets in the coordinate system of the other dataset.
\item[Pancakeplot:] This is the ``classical'' PCADSC.
\end{description}

Below we describe the details of these plots. Consider two different datasets in the same $d$ variables. Let $\lambda_{k1} \ge \dotsm \ge \lambda_{kd}$ and $\eta_{k1},\dotsc,\eta_{kd}$ be the variation components and the loadings for dataset $k=1,2$.

\medskip

\textbf{CumVarAgreement:} To be done.

\bigskip

\textbf{Hairplot:} Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets. The empirical covariance matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
\begin{equation*}
S_1 = \sum_{k=1}^d \lambda_{1j} \eta_{1k} \eta_{1k}^\top
= \lambda_{\max} \sum_{k=1}^d 
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg) 
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top.
\end{equation*}
We propose to visualize this decomposition by plotting arrows with length $l_{kj}$ and angle $\theta_{kj}/2$ anticlockwise from the diagonal in the $j$th row and the $k$th column of a $d \times d$ grid display, where
\begin{align*}
l_{kj} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k} \eta_{2j}^\top \rvert, &
\theta_{kj} &= \acos(\lvert \eta_{1k} \eta_{2j}^\top \rvert).
\end{align*}
Using the same grid display, but interchanging the first and the second dataset as well as the rows and the columns, and using the angles clockwise, similar arrows are plotted in a different color.

In our implementation we have used blue for the first and the second set of arrows, respectively. In this plot the blue arrows in the $k$th column visualize the decomposition of the $k$th principal components for the first dataset in the coordinate system of the second dataset. The scaling of the length by $\lambda_{\max}$ is made so that the longest arrow has unit length, and the absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy also implies the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$. We can visualize both the decomposition of $S_1$ and $S_2$ in the same plot by divide the angles by two and using anticlockwise and clockwise shifts from the diagonal. Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components for the second dataset in the coordinate system of the first dataset.

If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as the off-diagonal arrows would have zero length. Differences in the variation components are visualized as differences in the lengths of the arrows, also in the diagonal. And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal as well as arrows of non vanishing length in the off-diagonal.

\bigskip

Remarks: Presently permutation tests based on the Kolmogorov-Smirnov and the Cram\'er-von Mises test statistics has been implemented for CumVarAgreement. This should also be possible for the Hairplot based on the off-diagonal components. For the Pancakeplot a Wallyplot has been implemented (however, in Claus' Wally plot the position of the ``observed dataset'' is also random).

\bigskip

}

@@

\begin{figure}
\center
\includegraphics[scale = 0.8]{essWallyPCADSC.pdf}
\caption{\hl{something.}}
\label{plotESSPCADSCWally}
\end{figure}

\begin{figure}
\center
\includegraphics[scale = 0.8]{essHairplot.png}
\caption{The blue arrows show the principal components of the first dataset decomposed in the coordinate system of the principal components of the second dataset. And the red arrows the reverse.}
\label{plotESSHairplot}
\end{figure}

\begin{figure}
\center
\includegraphics[scale = 0.8]{essCumVarAgreement.png}
\caption{\hl{something.}}
\label{plotCumVarAgreement}
\end{figure}

@@

{\color{red} 
Figure \ref{plotESSHairplot} shows the hairplot. The blue arrows visualize the decomposition of the principal components for the first dataset in the coordinate system of the second dataset. We see that PC2 also loads on PC3, that PC3 also loads on PC4, and that PC4 also loads on PC2 and PC3. The red arrows visualize the decomposition of the principal components for the second dataset in the coordinate system of the first dataset. We see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3. For PC1, PC5 and PC6 the main difference is in the size of the variation component.
}
